Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-12-19 paper/paper_no_event.tex
Paper: fix structure of PP terms in appendix.
Made Michael's suggested changes to paper.

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a frequency-domain surrogate version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides. The version presented here is improved by the inclusion of the spin-induced quadrupole moment effect, and completed by a prescription for tapering the end of the waveform to reproduce qualitatively numerical simulations. The resulting model has 14 intrinsic parameters, so we reduce its dimensionality by using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. The previous implementation of this model could take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations. We therefore construct a fast and faithful frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz.

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can measure BNS properties such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}. Furthermore, an estimate of the BNS merger rate of 320--4740 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{GW170817} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be stacked to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.

The reliability of these measurements, however, depend on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Most time-domain waveform implementations involve solving a set of ordinary differential equations, which depending on the model can take seconds to hours for long BNS waveforms. The results then need to be transformed into the frequency domain where the analysis takes place. Because of this, analytic approximations providing waveforms directly in the frequency-domain  are often used instead.

For the analysis of GW170817 in Ref.~\cite{BNSPE}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

@@

Although these EOB models are more expensive than frequency domain models, parameter estimation results are still possible. In Ref.~\cite{BNSPE}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems. It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters. As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

@@

This work also takes a different approach from the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{uncertaintysampling.pdf}
\caption{Estimated $1\sigma$ RMS phase error $\epsilon_{\rm RMS}(\bx)$ (Eq.~\eqref{eq:errorrms}) at the frequency nodes $MF_j$ maximized over parameter space $\bx$ as a function of the number of samples. The black curve represents the estimated error using the Mat\'{e}rn kernel, and the gray curve represents the estimated error using the squared exponential kernel. Also shown are box plots of the maximum RMS phase errors at the nodes $MF_j \le 0.03$ for the surrogate compared to the test set of 1000 waveforms (Sec.~\ref{sec:accuracy}). The box plot represents the 25\%, 50\%, and 75\% quartiles. The whiskers contain all samples within 1.5 times the interquartile range. The black dots are outliers. Blue box plots: phase error from surrogates constructed using the initial design (160 corner and LHD waveforms) using either the Mat\'{e}rn or squared exponential kernel. Green box plots: surrogates constructed with the 160 initial waveforms and 400 uniformly sampled waveforms. Red box plots: surrogates constructed with the 160 initial waveforms and 400 waveforms chosen with uncertainty sampling. Magenta box plots: surrogates constructed with the 960 waveforms combining all sampling methods.}
\label{fig:uncsamp}
\end{figure}

@@

As discussed in more detail in~\cite{BNSPE}, the posterior $p(\vec\vartheta | d)$ for the parameters $\vec\vartheta$ given the detector data $\vec d$ is given by Bayes' theorem
\begin{equation}
p(\vec\vartheta | \vec d) \propto p(\vec\vartheta) \mathcal{L}(\vec d | \vec\vartheta),
\end{equation}
where $p(\vec\vartheta)$ is the prior and $\mathcal{L}(\vec d | \vec\vartheta)$ is the likelihood. 
For the prior we choose the sky position and orientation to be uniform on the unit sphere; the distance to be uniform in co-moving volume; the masses, spins, and tidal parameters to be uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$; and the mass ratio is restricted to the range $q\in[0.5, 1]$. 
As was done in~\cite{BNSPE}, we sample the data and waveform at 4096~Hz such that the Nyquist frequency is 2048~Hz, and integrate the likelihood function in the interval $[20, 2048]$~Hz. We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires approximately $10^7$ iterations to produce 10,000--20,000 statistically independent samples.

@@

The point particle terms coefficients up to 3.5PN can be found e.g. in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}. The non-zero coefficients read
\begin{align}
	a_{2} &= \frac{55 \eta }{9}+\frac{3715}{756} \nonumber\\
	a_{3} &= -16 \pi \nonumber\\
	a_{4} &= \frac{3085 \eta ^2}{72}+\frac{27145 \eta }{504}+\frac{15293365}{508032} \nonumber\\
	a_{5} &= -\frac{65 \pi  \eta }{9}-\frac{65}{3} \pi  \eta   \ln v +\frac{38645}{252} \pi   \ln v +\frac{38645 \pi }{756} \nonumber\\
	a_{6} &= -\frac{127825 \eta ^3}{1296}+\frac{76055 \eta ^2}{1728}+\frac{2255 \pi ^2 \eta }{12} \nonumber\\ 
	& -\frac{15737765635 \eta }{3048192}-\frac{6848  \ln v }{21}-\frac{640 \pi ^2}{3}-\frac{6848 \gamma_{E} }{21} \nonumber\\ 
	& +\frac{11583231236531}{4694215680}-\frac{13696 \log (2)}{21} \nonumber\\
	a_{7} &= -\frac{74045 \pi  \eta ^2}{756}+\frac{378515 \pi  \eta }{1512}+\frac{77096675 \pi }{254016}
\end{align}

>>>>

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a frequency-domain surrogate version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides. The version presented here is improved by the inclusion of the spin-induced quadrupole moment effect, and completed by a prescription for tapering the end of the waveform to qualitatively reproduce numerical relativity simulations. The resulting model has 14 intrinsic parameters. We reduce its dimensionality by using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. The implementation of the time-domain model can take up to an hour to evaluate using a starting frequency of 20~Hz, and this is too slow for many parameter estimation codes that require $\mathcal{O}(10^7)$ sequential waveform evaluations. We therefore construct a fast and faithful frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of $4.5\times 10^{-4}$ for the Advanced LIGO detector, and requires $\sim 0.13$~s to evaluate for a waveform with a starting frequency of 20~Hz. Finally, we perform an end-to-end test of the surrogate with a set of parameter estimation runs, and find that the surrogate accurately recovers the parameters of injected waveforms.

@@

The detection of the merging binary neutron star (BNS) system GW170817~\cite{GW170817} has demonstrated that the Advanced LIGO~\cite{Harry2010} and Advanced Virgo~\cite{Acernese2009} gravitational-wave detectors can measure BNS properties such as the masses, spins, and tidal parameters~\cite{BNSPE}, and these results can even be extended to measure quantities such as the radii and equation of state (EOS) of NSs~\cite{RaithelOzelPsaltis2018, DeFinstadLattimer2018, EOSPaper}. Furthermore, an estimate of the BNS merger rate of 110--3840 events Gpc$^{-3}$~yr$^{-1}$ (90\% confidence interval)~\cite{O2Catalog} from this event indicates that, when second generation detectors reach design sensitivity, we may eventually observe a population of tens or hundreds of events after a couple of years of observation~\cite{LIGORate2010}, and KAGRA~\cite{Somiya2012} and LIGO-India~\cite{IyerSouradeepUnnikrishnan2011} could improve this further. These multiple events could then be ``stacked'' to significantly improve measurements of the EOS~\cite{DelPozzoLiAgathos2013, LackeyWade2015}, and also measure the population distribution of masses and spins.

The reliability of these measurements, however, depends on Bayesian analyses that require fast and accurate waveform models to match to the gravitational-wave data. Standard Bayesian parameter estimation tools such as Markov chain Monte Carlo (MCMC) and Nested Sampling often require $\mathcal{O}(10^7)$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. To avoid large systematic errors, highly accurate waveform models are needed that include all relevant physical effects such as point-particle interactions, spins, and tides~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Most time-domain waveform implementations involve solving a set of ordinary differential equations, which depending on the model can take seconds to hours for long BNS waveforms. The results then need to be transformed into the frequency domain where the analysis takes place. Because of this, analytic approximations providing waveforms directly in the frequency-domain  are often used instead.

For the analysis of GW170817 in Refs.~\cite{BNSPE, O2Catalog}, therefore, a set of four frequency-domain waveform models were used as templates for the main results. These models are modifications to spinning binary black hole (BBH) waveform models (see Table I of Ref.~\cite{BNSPE} and references therein). They use either an aligned-spin post-Newtonian (PN) approximant in the Fourier domain (TaylorF2), the aligned-spin effective one body (EOB) formalism (SEOBNRT) or the phenomenological formalism for aligned-spin (PhenomDNRT) and precessing spin (PhenomPNRT). Tidal interactions were then included by simply adding a correction to the phase of each waveform. For the TaylorF2 waveform, the analytic 5PN and 6PN order terms were added. For the other three models, a fit that combined results from PN and numerical BNS simulations (referred to as the NRTidal fit in Ref.~\cite{BNSPE}) was added~\cite{DietrichBernuzziTichy2017}.

@@

Although these EOB models are more expensive than frequency-domain models, parameter estimation results are still possible. In Refs.~\cite{BNSPE, O2Catalog}, an alternative parallelized code, \texttt{RapidPE}, was also used that approximately fits the posterior density function (PDF) for the intrinsic waveform parameters instead of directly sampling it~\cite{PankowBradyOchsner2015, LangeOshaughnessyRizzo2018}. 
%Unfortunately, direct comparisons of the PDF between this code and traditional samplers using the same waveform models have not been published for BNS systems. It is therefore not clear how accurate this code is for measuring quantities such as the tidal parameters. 
The measured tidal parameter using EOB waveforms and \texttt{RapidPE} were broadly consistent with the results using frequency-domain waveforms and traditional samplers (Fig. 9 of Ref.~\cite{O2Catalog}). However, a direct comparison using identical waveforms has not yet been published for BNS systems.
As another possible solution for doing parameter estimation with EOB models, significant work has been done to optimize some EOB implementations, and speedups of a factor of several hundred~\cite{DevineEtienneMcWilliams2016} have been achieved. Different formulations of the differential equations can also speed up integration~\cite{NagarNettegno2018}. However, many of these optimizations are specific to each waveform model, and it is not clear that sufficient speed ups can be achieved for the \texttt{SEOBNRv4T} model discussed here with dynamical tides. As an alternative to these methods, we will develop a surrogate model of \texttt{SEOBNRv4T} that bypasses the need for specialized parameter estimation tools and optimization techniques.

@@

This work also takes a different approach from the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas}, multi-banding~\cite{VinciguerraVeitchMandel2017}, and relative binning~\cite{Zackay:2018qdy}.

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.49\textwidth]{uncertaintysampling.pdf}
\caption{Estimated $1\sigma$ RMS phase error $\epsilon_{\rm RMS}(\bx)$ (Eq.~\eqref{eq:errorrms}) at the frequency nodes $MF_j$ maximized over parameter space $\bx$ as a function of the number of samples. The black curve represents the estimated error using the Mat\'{e}rn kernel, and the gray curve represents the estimated error using the squared exponential kernel. 400 samples were added to the initial 160 samples for a total of 560 samples. Also shown are box plots of the maximum RMS phase errors at the nodes $MF_j \le 0.03$ for the surrogate compared to the test set of 1000 waveforms (Sec.~\ref{sec:accuracy}). The box plot represents the 25\%, 50\%, and 75\% quartiles. The whiskers contain all samples within 1.5 times the interquartile range. The black dots are outliers. Blue box plots: phase error from surrogates constructed using the initial design (160 corner and LHD waveforms) using either the Mat\'{e}rn or squared exponential kernel. Green box plots: surrogates constructed with the 160 initial waveforms and 400 uniformly sampled waveforms. Red box plots: surrogates constructed with the 160 initial waveforms and 400 waveforms chosen with uncertainty sampling. Magenta box plots: surrogates constructed with the 960 waveforms combining all sampling methods.}
\label{fig:uncsamp}
\end{figure}

@@

As discussed in more detail in~\cite{BNSPE}, the posterior $p(\vec\vartheta | d)$ for the parameters $\vec\vartheta$ given the detector data $\vec d$ is given by Bayes' theorem
\begin{equation}
p(\vec\vartheta | \vec d) \propto p(\vec\vartheta) \mathcal{L}(\vec d | \vec\vartheta),
\end{equation}
where $p(\vec\vartheta)$ is the prior and $\mathcal{L}(\vec d | \vec\vartheta)$ is the likelihood. 
For the prior we choose the sky position and orientation to be uniform on the unit sphere; the distance to be uniform in co-moving volume; the masses, spins, and tidal parameters to be uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$; and the mass ratio is restricted to the range $q\in[0.5, 1]$. 
As was done in~\cite{BNSPE}, we sample the data and waveform at 4096~Hz such that the Nyquist frequency is 2048~Hz, and integrate the likelihood function in the interval $[20, 2048]$~Hz. We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal} which requires $\sim 10^7$ iterations to produce 10,000--20,000 statistically independent samples.

@@

The point particle terms coefficients up to 3.5PN can be found e.g. in Eq.~(3.18) of Ref.~\cite{BuonannoIyerOchsner2009}, and have the structure
\be
	\Phi_{\rm F2}^{\rm PP}(\eta) = \sum_{k=0}^{7} a_{k} v^{k} + \mathcal{O}(v^{8}) \,.
\ee
The non-zero coefficients read
\begin{align}
	a_{0} &= 1 \nonumber\\
	a_{2} &= \frac{55 \eta }{9}+\frac{3715}{756} \nonumber\\
	a_{3} &= -16 \pi \nonumber\\
	a_{4} &= \frac{3085 \eta ^2}{72}+\frac{27145 \eta }{504}+\frac{15293365}{508032} \nonumber\\
	a_{5} &= -\frac{65 \pi  \eta }{9}-\frac{65}{3} \pi  \eta   \ln v +\frac{38645}{252} \pi   \ln v +\frac{38645 \pi }{756} \nonumber\\
	a_{6} &= -\frac{127825 \eta ^3}{1296}+\frac{76055 \eta ^2}{1728}+\frac{2255 \pi ^2 \eta }{12} \nonumber\\ 
	& -\frac{15737765635 \eta }{3048192}-\frac{6848  \ln v }{21}-\frac{640 \pi ^2}{3}-\frac{6848 \gamma_{E} }{21} \nonumber\\ 
	& +\frac{11583231236531}{4694215680}-\frac{13696 \log (2)}{21} \nonumber\\
	a_{7} &= -\frac{74045 \pi  \eta ^2}{756}+\frac{378515 \pi  \eta }{1512}+\frac{77096675 \pi }{254016} \,,
\end{align}
where $\gamma_{E}$ is Euler's constant.
